Iterative Monomorphisation

### 2024-05-02 iterative_monomorphisation.tex
edits together

The monomorphisation algorithm is paramatrised by many bounds, the first part of the evaluation process is to find appropriate values for these bounds. The second part compares the performance of Zipperposition with and without monomorphisation. The third part tests the relative performance of different provers on polymorphic problems and their monomorphised counterparts.

The test problems are taken from the TPTP library. The TPTP library contains 1765 problems in the TF1 and TH1 grammars, they are the first-order and higher-order problems with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems including real numbers or importing axioms including real numbers. In total our testing suite contains 1754 polymorphic problems. We chose as a measure of success the number of problems that could each be solved by an instance of a prover in under 30 seconds with a single thread. This setup mirrors the setting of interactive provers in which an instance of an automatic prover will typically be called for similar durations.

@@

As a precaution against overfitting, all tests related to parameter optimisation were made on 500 randomly chosen problems out of 1754. Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options with which Zipperposition will be instantiated. Zipperposition has a portfolio mode with several different configurations, we evaluated all base configurations which called E against our test set of 500 problems without using monomorphisation. The best performing configuration became the basis upon which we tested the different monomorphisation options. Our evaluation found that the ``40 e lift'' configuration performed best by solving 136 problems in under 30 seconds each.%[TODO recheck this]

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time. Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed. Given the number of bounds however, this is not feasible. Instead, we group closely related bounds together and test combinations of values for bounds in these groups. Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group. If several values for a given bound solve the same number of problems, the values solving the greater number of problems with monomorphisation will be selected, if these numbers are also equal, the most constraining bounds will be selected. This method requires an initial default set of bounds and reasonable ranges of values for each bound. We determined these values based on the results of preliminary evaluations.

Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration. The limit of newly generated type arguments per clause is determined by three components which form a natural group of bounds. Our results show that generating no new monomorphic type arguments seems to be the best approach. It is possible to monorphise formuale without generating monomorphic type arguments because non-monomorphic type argument generation can generate substitutions which instantiate one or more type variables and it is these substitutions which are used to monomorphise formulae. This table also illustrates the impact that the choice of bounds can have on performance. The choice values of only three bounds can result in an increase of up to 30\% in the number of problems solved.

@@

Table~\ref{nmon_ty_args} fills the same role as table~\ref{mono_ty_args} except it is for bounds limiting the number of new non-monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations. These results confirm that non-monomorphic argument generation drive the creation of useful non-monomorphic clauses. This is indicatied by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems largely unaffected by bounds' values when within a certain range but drops off significantly beyond this range.
%TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)

>>>>

The monomorphisation algorithm is parametrised by many bounds. The first part of the evaluation process is to find appropriate values for these bounds. The second part compares the performance of Zipperposition without and with monomorphisation. The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.

The benchmarks are taken from the TPTP library \cite{tptp}. The TPTP library version TODO contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include real numbers. In total, our benchmark suite contains 1754 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in under 30 seconds with a single thread. This corresponds to the default time limit in Sledgehammer.

@@

As a precaution against overfitting, the part of the evaluation related to parameter optimisation was carried out on 500 randomly chosen problems out of the 1754 selected problems, and the rest of the evaluation was carried out on the remaining 1254 problems. Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition. Zipperposition has a portfolio mode with several configurations. In a preliminary experiment, we informally evaluated all portfolio configurations that called E against our benchmark suite of 500 problems without using monomorphisation. The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options. Our informal evaluation found that the so-called \verb|40_e_lift| configuration performed best by proving 136 problems.%[TODO recheck this]

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time. Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed. Given the number of bounds, however, this is not feasible. Instead, we group closely related bounds together and test combinations of values for bounds in these groups. Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected,
and shown in bold in Tables \ref{mono_ty_args} to \ref{xxx}. This method requires an initial default set of bounds and reasonable ranges of values for each bound. We determined these values based on the results of preliminary experiments on the benchmark suite of 500 problems.

Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration. The limit on newly generated type arguments is determined by three components that form a natural group of bounds. The table shows that generating no new monomorphic type arguments seems to be the best approach. This result may seem counterintuitive, but it is possible to monomorphise formulae without generating monomorphic type arguments. This is because non-monomorphic type argument generation can generate substitutions that instantiate one or more type variables, and it is these substitutions that are used to monomorphise formulae.

@@

Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems largely unaffected by the bounds' values when within a certain range but drops off significantly beyond this range.
%TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)
